---
title: "gbif_datacleaning"
author: "P Bal"
collaborators: "Casey Visintin, Skip Woolley"
date: "`r Sys.Date()`"
output: html_document
R script file: "gbif-data_processing.R"
gbif data citation: "GBIF.org (12th December 2017) GBIF Occurrence Download https://doi.org/10.15468/dl.kaciwi"
---

<!-- Specify global options for code chunks -->
```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = FALSE)
```



## Load packages
```{r load packages, eval=TRUE, echo=TRUE}
library(pacman)
p_load(DBI, RPostgreSQL, data.table, rgbif, foreach, tidyverse, foreach, iterators, parallel, doParallel, doMC) 
  # tidyverse includes dplyr so no need to load that separately
```



## Connect to server
<span style="color:blue">Issue: Prompt for server details here</span>
```{r connect to server, eval=FALSE, echo=TRUE}
drv <- dbDriver("PostgreSQL")  #Specify a driver for postgreSQL type database
con <- dbConnect(drv, dbname="", user="", password="", host="", port="")
```



## GBIF record counts
*Number of records in the online gbif database as of* `r Sys.Date()`
```{r count gbif all, eval=TRUE, echo=FALSE}
occ_count()
```
Ref: https://www.gbif.org/occurrence/search
 
*Number of records in the downloaded gbif csv file downloaded on 2017-12-12*<br/>
  - i.e. number of rows in downloaded csv file named gbif.raw_data on server<br/>
  - Count obtained from db metadata 
```{r count gbif.raw_data, eval=TRUE, echo=FALSE}
raw.n <- as.numeric(dbGetQuery(con,"SELECT reltuples::bigint AS estimate
                FROM   pg_class
                WHERE  oid = 'gbif.raw_data'::regclass;
                "))
raw.n
```

*Size of gbif database (see citation above)*
```{r sizeraw, eval=TRUE, echo=TRUE}
dbGetQuery(con,"
              SELECT pg_size_pretty(pg_relation_size('gbif.raw_data'));
              ")
```


## Run first filter on downloaded gbif csv data and create new table
  - raw data: gbif.raw_data
  - new table: gbif.firstfilter
  - filter includes: (1) Kingdom = Animalia; (2) Geographical coordinates = NOT NULL; and (3) Time frame = 1950 - 2017

```{r ffilter, eval=FALSE, echo=TRUE}
## Run query for first filter and create new table
dbSendQuery(con,"
           CREATE TABLE gbif.firstfilter AS
           SELECT *
           FROM gbif.raw_data
           WHERE kingdom = 'Animalia'
           AND decimallatitude IS NOT NULL
           AND decimallongitude IS NOT NULL
           AND year BETWEEN '1950' AND '2017';
           
           ALTER TABLE gbif.firstfilter 
           DROP COLUMN catalognumber,
           DROP COLUMN collectioncode,
           DROP COLUMN identifiedby,
           DROP COLUMN institutioncode,
           DROP COLUMN mediatype,
           DROP COLUMN publishingorgkey,
           DROP COLUMN recordedby,
           DROP COLUMN rightsholder,
           DROP COLUMN typestatus;
           
           ALTER TABLE gbif.firstfilter ADD PRIMARY KEY (gbifid);
           
           ALTER TABLE gbif.firstfilter 
           ALTER COLUMN gbifid TYPE INTEGER USING gbifid::integer;
           
           CREATE INDEX firstfilter_index
           ON gbif.firstfilter USING btree
           (taxonkey COLLATE pg_catalog.default)
           TABLESPACE pg_default;
           ") 
```

*Number of records in new table after running first filter* <br/>
  - i.e. number of rows in gbif.first_filter<br/>
  - Count obtained from db metadata
```{r count gbif.firstfilter, eval=TRUE, echo=FALSE}
firstfilter.n <- as.numeric(dbGetQuery(con,"
                SELECT reltuples::bigint AS estimate
                FROM   pg_class
                WHERE  oid = 'gbif.firstfilter'::regclass;
                "))
firstfilter.n
```


## Change data type within table
#### Convert empty characters in new table to NULL
<span style="color:blue">Issue: Prompt for server details here</span>

```{r nullval, eval=FALSE, echo=TRUE}
## Obtain field names for gbif.firstfilter
gbif.fields <- read.csv(paste(getwd(), "/R/gbif_header.csv", sep=""),sep="\t")
gbif.fields <- colnames(gbif.fields)
drop.fields <- c("catalognumber", "collectioncode", "identifiedby", "institutioncode", "mediatype", "publishingorgkey", "recordedby", "rightsholder", "typestatus", "order")
# "order" is a reserved word in PostgreSQL because it is an operator (e.g. ORDER BY x). Unfortunately, GBIF also uses this as a field name. So it doesn't work in the foreach loop. Instead, it needs to be run as separate query. 
in.fields <- gbif.fields[which(!gbif.fields %in% drop.fields)]

## Specify empty characters in all columns within gbif.firstfilter (except order) as NULL
p_load(foreach, iterators, parallel, doParallel, doMC)

registerDoMC(detectCores() - 1) #set up cores for parallel processing (72-1)

foreach (i=in.fields) %dopar% { 
  drv <- dbDriver("PostgreSQL")  #Specify a driver for postgreSQL type database
  con <- dbConnect(drv, dbname="", user="", password="", host="", port="")
  dbSendQuery(con, paste("UPDATE gbif.firstfilter SET ",i," = NULL WHERE ",i,"::char = '';", sep=""))
}

# Set empty characters in the gbif field 'order' to NULL
dbSendQuery(con, "UPDATE gbif.firstfilter SET \"order\" = NULL WHERE \"order\"::char = '';")
```


#### Convert lat-long to numeric
```{r latlong numeric, eval=FALSE, echo=TRUE}
dbSendQuery(con,"
                ALTER TABLE gbif.firstfilter 
                ALTER COLUMN decimallatitude TYPE NUMERIC USING decimallatitude::numeric
                ")

dbSendQuery(con,"
                ALTER TABLE gbif.firstfilter 
                ALTER COLUMN decimallongitude TYPE NUMERIC USING decimallongitude::numeric
                ")
```

## Run checks on new table
*Size of new table*
```{r sizeff, eval=TRUE, echo=FALSE}
dbGetQuery(con,"
              SELECT pg_size_pretty(pg_relation_size('gbif.firstfilter'));
              ")
```

*Number of columns*
```{r colcount, eval=TRUE, echo=FALSE}
dbGetQuery(con,"
            SELECT COUNT(*) from information_schema.columns
            where table_name='firstfilter';
            ")
```

*Check for NULL values in table...??*
...





## Run second filter
  - on gbif.first_filter to create gbif.second_filter
  - includes the following steps:
    1. Issues
    2. Accuracy
    3. Basis of Record
    
    
```{r filter2 step1}

```


```{r filter2 step2}

```


## References and useful links
*gbif data fields* <br/>
https://www.gbif.org/developer/occurrence <br/>
http://rs.gbif.org/core/dwc_occurrence.xml

*knitr* <br/>
http://kbroman.org/knitr_knutshell/pages/Rmarkdown.html
